---
title: 'Project 2: Group 1 - Boston Housing Racial Bias'
author: "Will Johnson"
date: "12/3/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

# Exploring our Dataset

```{r}
# Download our dataset
library(mlbench)
data(BostonHousing2)
data <- BostonHousing2
head(data)
```



## "Majority Black" response variable

The variable "b" is described in the census documents as...

> 1000(B - 0.63)^2 where B is the proportion of blacks by town

Originally, we had planned to set this as a “majority black” binary variable. Based on the description, we thought the values would be between 0 and 1, or possibly 0 and 100. But the variable itself was a strange one instead. Take a look at the distribution of this variable `b`. 

```{r}
# plot the distribution on the "b" variable. 
ggplot(data, aes(x = b))+
  geom_histogram(binwidth = 1)
```
Not only is the range very odd (0 to 400), the majority of these towns are in the high-end, suggesting that the data has been transformed so that the higher the value is, the whiter the neighborhood.

The most interesting part about our findings is that if we try to reverse engineer the original B value (based on their seemingly arbitrary formula b=1000(B−.63)2), it’s not possible to get the answer. The result of the previously-mentioned formula is a quadratic curve, plotted below:

```{r}
# Using the dataset's documentation, plot potential values for proportions of black people
# against each value of `data$b`. 
props <- c()
i <- .01
for(x in 1:99){
    props <- c(props,1000*(i-.63)**2 )
    i = i + .01
}
parab <- data.frame(props,seq(.01, .99, by=.01))
colnames(parab) = c('x','y')
ggplot(data = parab, aes(x=y, y=x ))+
  geom_line()+
  labs(x = "Potential proportion of black people", y='b variable')

```

Looking at this, it appears that the dataset is impossible to perfectly break back down into its original B variable because for some values, there are potentially 2 values of B for one value of the transformed b variable. Specifically, anything more than 26% black in a neighborhood becomes vague.


Some brief external research suggests that this transformation on the variable was set to create a pseudo-parabolic relationship to account for an initial drop in value when the proportion was too mixed-race. After 75% black, the effect was expected to rise again because of preference for a neighborhood of one’s own race or “self-segregation” [Harrison,Rubinfeld, 96].

To quote an assistant professor of Mathematics at CUNY who researched this:

> Harrison and Rubinfeld appear to have decided on a threshold of 63% at which to switch the regime of price decline to price increase (i.e. a so-called “ghetto threshold”) [Carlisle,1]


Given that this was their decision at the time, we’ll follow the precedent they set in the 1970’s. We’ll assess if there is a relationship between where a home falls on that threshold to see its effect on the price of a home, as well as try to predict where on that threshold a house will fall for our logistic model.

Plugging our 26% threshold into that formula, we get:

$1000*(.26−.63)2=136.9$

so black_neighborhood = 1 if b <= 136.9

```{r}
# code to create binary black_neighborhood variable
data <- data %>% 
           mutate(black_neighborhood = (b <= 136.9) * 1)
# what % of our neighborhoods are black neighborhoods by this definition? 
sum(data$black_neighborhood)/nrow(data)
```

Finally, let’s chart this variable to see how some of the other’s fall across it. Namely our price variable

```{r}
data$black_neighborhood <- as.factor(data$black_neighborhood)
# plot our new variable and the price distributions across it. 
ggplot(data, aes(x= black_neighborhood, y=medv))+
  geom_boxplot(fill="light blue")

```


Definitely seems like there’s a relationship between these two and price. Looking at it, we should assume there should be predictive value including this variable in predicting price.


# Part 1 -Regression : How much does the relative blackness of a neighborhood affect price? 

First we'll start by looking at a handful of other visuals to see which variables seem to affect price beyond our "black_neighborhood" variable. We'll start with looking out our prices overall.

```{r}
ggplot(data, aes(x=medv))+
  geom_histogram()+
  labs(title = "Distribution of Median Housing Values per Area")
```


Looks mostly normally distributed with a handful of towns with a particularly high median value. We should look out for those to be potential outliers later. 

Next, some scatterplots between quantitative variables. 



```{r}
ggplot(data, aes(x = log(crim), y = medv))+
  geom_point(color = 'green3')+
  labs(title="Crime Rate (log) by Median Home Value",
       x = "Crime Rate - Log Scale")
```


Vaguely, it looks as though the more the crime rate increases, the values of houses start to decrease. We used a log scale for crime here because otherwise the majority of towns had too low a crime rate to see this trend. 

Next,

```{r}
head(data)
```

